---
title: "Correspondence Analysis"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
```

## Goodness of Fit

* All the above tests are called goodness-of-fit tests
* They quantify how well a model fits observations
* They measure discrepancy between observed values and expected values under the model
* Can we visualize the residuals?
* Rather than reject this global null hypthesis, can we find what is driving the statistic?
* Two ways:
    * Visualization of contingency table with association and mosaic plots
    * Visualization of deviation from model with correspondence analysis

## Contingency Tables

* Several discrete random variables with same categories $1,\dots,c$
    * Test of homogeneity
    * Compare distribution of all random variables
    * Are the observation generated from one multinomial distribution with one probability vector
    * Possibly different number of observations
    * Experimental design: TODO
* Two discrete random variables with different categories $1,\dots,r$ and $1,\dots,c$
    * Test of independence
    * Same number of observations
    * Experimental design: TODO

## Association Plots

* This is a followup on $\chi^2$ tests when they are rejected
* Pearson residuals
$$r_{ij} = \frac{O_{ij}-E_{ij}}{\sqrt{E_{ij}}}$$
* and the $\chi^2$ statistics is squaring and summing over all cells
$$\chi^2 = \sum_j \sum_i r_{ij}^2$$

## Association Plots

* Each cell of contingency table is represented by a rectangle encoding information in width, height, location, and color of the rectangle
* *Height:* Proportional to Pearson residual $\frac{O_{ij}-E_{ij}}{\sqrt{E_{ij}}}$
* *Width:*  Proportional to $\sqrt{E_{ij}}$
* *Area:* Proportional to $O_{ij}-E_{ij}$
* *Baseline:* 
    * If $O_{ij} > E_{ij}$, then rectanble above
    * Otherwise rectangle below
* *Color:* Standardized Pearson residuals that are asymptotically standard normal

## Association Plots Example

Distribution of alcoholic status same for different type of crime?

Contingency table with frequencies of  
criminals who committed crimes (6 RV's) and  
their alcoholics status (category: alcoholic and non-alcoholic)

```{r}
c1 = c(50,88,155,379,18,63) 
c2 = c(43,62,110,300,14,144)
ct = cbind(c1,c2)
colnames(ct) = c("Alcoholic","Non-Alcoholic")
rownames(ct) = c("Arson","Rape","Violence","Theft","Coining","Fraud")
ct
```

## Association Plots Example

```{r echo=TRUE, message=FALSE, fig.width=5.5, fig.height=5.5}
library(vcd); assoc(ct,shade = TRUE)
```

## Mosaic Plots

* *Width:* Relative frequencies of one variable
* *Height:* Relative  (conditional) frequencies of the second variable given the first
* *Color:* Standardized Pearson residuals that are asymptotically standard normal

## Mosaic Plots {.build}

Subdivide by marginal frequencies of first variable

```{r echo=TRUE}
frequency = rowSums(ct)
proportions = frequency/sum(frequency)
t(data.frame(frequency=round(frequency,0),proportions=round(proportions,2)))
```

Subdivide by marginal frequencies of second given first variable

```{r echo=TRUE}
t(ct/rowSums(ct))
```

## Mosaic Plots

```{r echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
mosaic(ct,shade = TRUE)
```

## Correspondence Analysis

* Actively developped in 1965 in France
* French school of data analysis: Geometry instead of probability
* Can be used to plot contingency tables for exploratory data analysis
* Decomposes the $\chi^2$ statistic into orthogonal factors
* To first approximation: extension Principal Components Analysis (PCA) by 
* replacing the variance in PCA with an inertia proportional to the $\chi^2$ distance of the table from independence
* decomposes this measure of departure from independence along axes that are orthogonal according to a $\chi^2$ inner product

## Correspondence Analysis

* Denote $m \times p$ contingency table as $N$
* Total number of observations $n = \sum_{i=1}^m \sum_{j=1}^p n_{ij}$
* Denote frequency table as $F = \frac{N}{n}$
* Vector of row sums of $F$: $\operatorname{diag}(D_r) = r = \frac{1}{n} N 1_p$
* Vector of column sums of $F$: $\operatorname{diag}(D_c) = c^T = \frac{1}{n} N^T 1_m$
* Peform Principle Component Analysis (PCA) on new data matrix
$$X = D_r^{-1} F D_c^{-1} − 1_m^T 1_p$$
* Generalized singular value decomposition on $X = USV^T$  
with constrains $U^T D_r U = I_m$ and $V^T D_c V = I_p$

## Correspondence Analysis

* Relationship to $\chi^2$ statistic
$$D_r X^T D_c X = \frac{\chi^2}{n}$$
* In standard PCA, principle components explain variance
* In CA, principle components explain devation from independence

## Mosaic, Association, and Correspondence Analysis Example

```{r echo=TRUE}
library(ca)
HairEye = margin.table(HairEyeColor, c(1, 2)); HairEye
```

----

```{r echo=TRUE}
chisq.test(HairEye)
```

----

```{r echo=TRUE, message=FALSE, fig.width=6.5, fig.height=6.5}
mosaic(HairEye, shade = TRUE)
```

---

```{r echo=TRUE, message=FALSE, fig.width=6, fig.height=6}
assoc(HairEye,shade = TRUE)
```

----

```{r echo=TRUE}
ca.res = ca(HairEye); ca.res
```

----

```{r fig.width=10, fig.height=5}
plot(ca.res)
```

Color code: <span style="color:blue">Hair</span>, <span style="color:red">Eye</span>
