---
title: "Logistics and Introduction"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Modern Statistics

To quote Andrew Gelman [(source)](http://arxiv.org/pdf/1001.2968.pdf):

```
          "If you wanted to do foundational research 
          in statistics in the mid-twentieth century, 
          you had to be bit of a mathematician, 
          ... if you want to do statistical research 
          at the turn of the twenty-first century, 
          you have to be a computer programmer."
```

## History of Nonparametric Statistics

Check on Google Ngram Viewer.

```{r warning = FALSE,echo = FALSE}
library(ggplot2)
library(ngramr)
ng = ngram(c("Nonparametric Statistics"), year_start = 1950)
ggplot(ng, aes(x=Year, y=Frequency, colour=Phrase)) + geom_line()
```

## Stats 205

* You can find everything on our course website: http://christofseiler.github.io/stats205/
* Homeworks every 8 days (send solutions to TA's)
* Office hours (check room availability): Wednesday 2:45 -- 3:45
* TA Office hours (write them an email): 2 x 1.5 hours

## Why?

1. Few assumptions about underlying populations from which data is obtained, e.g. populations don't need to follow a normal distribution
2. Often easier to understand and apply than parametric tests
3. Slightly less efficient than parametric test when parametric assumptions hold, but if assumptions don't hold then wildly more efficient
4. Jackknife and bootstrap can be used in many practical situations where theory is intractable
5. Bayesian methods are available so prior information can be incorporated

## Goals

* Get an overview of classical, Bayesian, and modern methods
* Learn how to implement methods yourself and use existing R packages
* Be aware and understand underlying assumptions
* Apply to modern data analysis problems that you care about

