---
title: "R and R Markdown Basics"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output: 
  ioslides_presentation: 
    smaller: yes
    transition: faster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## For This Course

During this course will write a lot of R code and run computer simulations.
All my examples and homeworks will be written in R. 
Choosing R for your homework solutions is highly recommended.

What is R:

- R is an interpreted programming language, which means you will not have to compile your code. 
- R is very interactive, which means that you can play around with vectors and matrices, plot results.
- To keep track of your progress and to be able to construction your analysis steps, we will use R markdown. 
- R markdown is a format to make web reports that you can share with your collaborators.

## Vectors

Make a vector:

```{r}
x = c(11,218,123,36,1001)
y = rep(1,5)
z = seq(1,5,by=1)
x + y
z + 10
```

## Operations on Vectors

Some operations:

```{r}
sum(y)
c(mean(z),sd(z))
length(z)
```

## Coin Tossing

And most importantly there is a lot of functions for statistics. 

Like tossing a coin three times:

```{r}
set.seed(1)
coin = c('H','T')
samples = sample(coin,100,replace = TRUE)
```

Should come about half of the times head:

```{r}
sum(samples == 'H')
```

## Matrices

We can combine vectors of the same type into matrices.

```{r}
X = cbind(x,y,z)
X
```

## Data Frames

Data frames are used to combind variables of different types in one object.

```{r}
subjects = c('Jim','Jack','Joe','Mary','Jean')
sex = c('M','M','M','F','F')
score = c(85,90,75,100,70)
D2 = data.frame(subjects,sex,score)
D2
```

## Generating Random Data

R provides a wide variaty of distributions that we can sample from. All the ones we know from intro stats courses. For instance the normal distribution:

```{r}
z = rnorm(1000)
mean(z)
sd(z)
```

## Basic Plotting

Plotting the sample from last slide.

```{r}
hist(z,breaks=30)
```

## Fancy Plotting

The ggplot2 package is very popular to make more sophisticated plots.

```{r}
library(ggplot2)
```

You will have to learn the grammar of ggplot. There are many tutorials online. Here is one example [link](http://www.cookbook-r.com/Graphs/). 

Let's see how it looks like in action on sleep study data.

```{r}
library(Lock5Data)
data(SleepStudy)
```

## Fancy Plotting

```{r eval=TRUE,echo=TRUE}
ggplot(SleepStudy, aes(x=Drinks,y=GPA)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) +
  geom_smooth() + xlab('number of alcoholic drinks per week')
```

## Repeated Tasks


