---
title: "Signed-Rank Wilcoxon Example"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output: 
  ioslides_presentation: 
    incremental: yes
    smaller: yes
    transition: faster
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question

Does sending kids to school improve social awareness over home schooling?

## Experiment

Eight pairs of identical twins who are of nursery school age.  
For each pair, one is randomly selected to attend nursery school while the other remains at home.  
At the end of the study period, all 16 children are given the same social awareness test.

```{r}
school = c(82,69,73,43,58,56,76,65)
home = c(63,42,74,37,51,43,80,62)
```

## Differences

For each pair, the response of interest is the difference in the twins’ scores:

```{r}
school
home
response = school - home
response
```

## Model

* $X_1,X_2,\dots,X_n$ denotes random sample following the model
$$ X_i = \theta + e_i, $$
random errors are independently and identically distributed with a continuous probability density function $f(t)$ symmetric around $0$.

* We test hypothesis:
$$ H_0: \theta = 0 \text{ versus } H_A: \theta > 0 $$
The null distribution of the test statistic does not depend on $f(t)$

## Ranks

Calculate ranks of absolute values:

```{r}
response
abs(response)
rank(abs(response))
```

## Statistic

We use signed-rank Wilcoxon procedure to find out.  
The usual statistic is the sum of the ranks of positive items:

```{r}
wplus = sum((response>0)*rank(abs(response)))
wplus
```

<!--
or equivalently:

```{r}
n = length(response)
w = sum(sign(response)*rank(abs(response)))
wplus = 0.5*w + n*(n+1)/4
wplus
```
-->

## Null Distribution of Test Statistic

Is this due to chance?  
Luckily, the probability mass function for our test statistic is known.  
To see this, let's look at an even simpler example of only three differences, with 8 possible outcomes:

```{r message=FALSE}
library(R.utils)
sign = intToBin(0:7)
sign = sapply(sign,function(i) c(substr(i,1,1),substr(i,2,2),substr(i,3,3)) )
sign = t(matrix(as.integer(sign),nrow = 3))
rank = t(matrix(rep(1:3,8),nrow = 3))
data.frame(sign=sign,rank=rank)
```

## Null Distribution of Test Statistic

```{r}
outcome = data.frame(sign=sign,rank=rank,mul=sign*rank,sum=apply(sign*rank,1,sum))
outcome
table(outcome$sum)
```

## Null Distribution of Test Statistic

Or through Monte Carlo simulations:

```{r}
n = 3; nsim = 10000
X = matrix(rnorm(n*nsim),ncol=n)
WplusSim = apply(X,1, function(x) { r = rank(abs(x)); sum(r * (x>0)) })
table(WplusSim)/nsim
```

and compare with theoretical result:

```{r echo=FALSE}
prop = c(1/8,1/8,1/8,2/8,1/8,1/8,1/8)
names(prop) = 0:6
prop
```

## Computations

Now back to our "more complicated" school example.

```{r eval=FALSE}
n = length(response)
x = seq(0,n*(n+1)/2,1)
colors = rep("white",length(x))
colors[x >= wplus] = "red"
barplot(dsignrank(x,n),
        names.arg = x,
        main = paste0("dsignrank(x, n = ", n, ")"),
        col = colors)
```

## Computations

```{r echo=FALSE}
n = length(response)
x = seq(0,n*(n+1)/2,1)
colors = rep("white",length(x))
colors[x >= wplus] = "red"
barplot(dsignrank(x,n),names.arg = x,main = paste0("dsignrank(x, n = ", n, ")"),col = colors)
```

## Computations

and using the cumulative distribution function $F_{W^+}(w^+-1,n)$ of our test statistic:

```{r}
plot(x,psignrank(x, n = n),type = "s",main = paste0("psignrank(x, n = ", n, ")"))
abline(v=wplus,col = "red"); abline(h=psignrank(wplus,n),col = "blue")
```

## Computations

we can find the $p$-value using $P_{H_0}(W^+ \ge w^+) = 1 - F_{W^+}(w^+ - 1,n)$:

```{r}
pvalue = 1-psignrank(wplus-1,n,lower.tail = TRUE)
pvalue
```

There is also an R base implementation of this test:

```{r}
wilcox.test(response,alternative="greater")
```

## Conclusion

We can reject the null hypothesis at significance level $\alpha = 0.05$ in favor of the alternative that sending kids to school improves their social awareness.
