---
title: "Homework 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation

1. Install RStudio: <http://www.rstudio.com>
2. If you are not familiar with R work thorugh this short online course: <http://tryr.codeschool.com/>
3. If you are not familiar with R markdown go throug this website: <http://rmarkdown.rstudio.com/>

The following exercise are taken from our textbook [(Stanford library link)](https://searchworks.stanford.edu/view/10718635).

## Exercises

### Exercise 1

Use the commands ``seq`` and ``rep`` create the following lists.

1. Even numbers less than 20
2. Odd numbers between 101 and 203
3. ``1 3 1 3 1 3 1 3``
4. ``1 1 1 1 3 3 3 3``

### Exercise 2

Calculate the mean and variance of the following. 

1. First 100 integers.
2. Random sample of 50 normal random variates with mean 30 and standard deviation 5.

### Exercise 3

Simulate the sampling distribution of the mean of 10 tosses of a fair die.

### Exercise 4

Approximate the power of a $t$-test of $H_0: \mu = 0$ versus $H_A: \mu > 0$ when the true mean is $\mu = 0.5$. Assume a random sample of size $n = 25$ from a normal distribution with $\sigma = 1$. Assume $\alpha = 0.05$.

### Exercise 5

Use the commands ``dnorm``, ``seq``, and ``lines`` to create a plot of the pdf of a normal distribution with $\mu = 50$ and 
$\sigma^2 = 10$.

### Exercise 6

Write an R function which computes the sign analysis. For example, the following commands compute the statistic S+, assuming that the sample is in the vector x.
```{r eval=FALSE}
xt <- x[x!=0]
nt <- length(xt)
ind <- rep(0,nt) 
ind[xt > 0] <-1
splus <- sum(ind)
```

### Exercise 7

Calculate the sign test for the nursery school example, Example 2.3.1. Show that the $p$-value for the one-sided sign test is $0.1445$.


## Deadline

Please email your solutions in the form of a Rmd file to Nan Bi and Lexi Guan.

This homework is due on Thursday, April 7th at 1:30 pm.

You are encouraged to work through this homework with your peers. But write your own code and explain the results in your own words. 
