---
title: "Time to Event Analysis"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  beamer_presentation:
    incremental: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
library(survival)
library(npsm)
```

## Overview

* Survival data
* Survival function
    * Kaplan-Meier estimator
    * Log Rank test
* Hazard function
    * Cox proportional hazards models

## Survival Data

* We are interested in time to an event of interest as the outcome variable
* In medicine:
    * Often in a clinical trial the goal is to evaluate the effectiveness of a new treatment at prolonging survival
    * For example to extend the time to the event of death
    * It is usually the case that at the end of followup a portion of the subjects in the trial have not experienced the event
    * For these subjects the outcome variable is **censored**

## Survival Data

* In engineering:
    * Similarly, in engineering studies, often the lifetimes of mechanical or electrical parts are of interest
    * In a typical experimental design, lifetimes of these parts are recorded along with covariates (including design variables)
    *   Often the lifetimes are called failure times, i.e., times until failure
    * As in a clinical study, at the end of the experiment, there may be parts which are still functioning (censored observations)
    
## Survival Data

* In this context, it one object of interesting is the survival function
* An popular estimator of this function is the Kaplan–Meier estimator
* It is common to estimate two function (e.g. in a case/control experiment) and to test the hypothesis that they are the same
* A popular test is the Log Rank Test
* An alternative test is the Gehan's Test
* Another object of interst is the hazard function
* Which can be interpreted as the instantaneous chance of the event (death)
* In this context, we'll talk about the Cox proportional hazards models

## Survival Function

* Let $T$ denote the time to an event
* Assume $T$ is a continuous random variable with cdf $F(t)$
* The survival function is defined as the probability that a subject survives until at least time $t$
$$S(t) = P(T > t) = 1 - F(t)$$
* When all subjects in the trial experience the event during the course of the study
* So that there are no censored observations, an estimate of $S(t)$ may be based on the empirical cdf
* However, in most studies there are a number of subjects who are not known to have experienced the outcome prior to the study completion
* Kaplan and Meier (1958) developed their product-limit estimate as an estimate of $S(t)$ which incorporates information from censored observations

## Survival Function (Example: No Censored Observations)

* Treatment of pulmonary metastasis (cancerous spreads to the lung) arising from bone cancer, survival time was collected
```{r echo=TRUE}
survTimes = c(11,13,13,13,13,13,14,14,15,15,17)
```
* No censored observation, we can estimate survival function at time *t* with empirical cdf
$$\widehat{S}(t) = \frac{\#\{t_i > t\}}{n}$$
* Estimate by counting
$$
\widehat{S}(t) = 
\begin{cases}
1 & 0 \le t < 11 \\
\frac{10}{11} & 11 \le t < 13 \\
\frac{5}{11} & 13 \le t < 14 \\
\frac{3}{11} & 14 \le t < 15 \\
\frac{1}{11} & 15 \le t < 17 \\
0 & t \ge 17
\end{cases}
$$

## Survival Function (Example: No Censored Observations)

```{r out.width=".8\\linewidth"}
t = seq(0,17,0.01)
Shat= sapply(t,function(t) sum(survTimes>t)/length(survTimes))
plot(t,Shat,type = "l",xlab = "Time (in months)",ylab = "Survival",
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
```

## Kaplan-Meier Estimator

* In most clinical studies, at the end of followup there are subjects who have yet to experience the event being studied
    * Either hasn't happened
    * Or the subject died before
* In such cases the Kaplan–Meier product limit estimate can be used
* Let $t(1) < \dots < t(k)$ denote the ordered distinct event times
* If there are censored responses, then $k < n$
* Let $ni = \#\text{subjects}$ at risk at the beginning of time $t(i)$ and $d_i = \#\text{events}$ occurring at time $t(i)$
* The **Kaplan-Meier estimate** of the survival function is defined as
$$\widehat{S}(t) = \prod_{t(i) \le t} \left(1- \frac{d_i}{n_i}\right)$$

## Kaplan-Meier Estimator

* TODO: Explain derivations with conditional probabilities

## Kaplan-Meier Estimator (Example)

* Data:
    * Relapse: 3 6.5 6.5 10 12 15
    * Lost to followup: 8.4
    * Alive and in remission at at end of study: 4 5.7 10
* Step-by-step Kaplan–Meier estimate:
$$
\begin{tabular}{r|r|r|r|r}
\hline
$t$ & $n$ & $d$ & $1-d/n$ & $S(t)$ \\
\hline
$3$ & $10$ & $1$ & $9/10=0.9$ & $0.9$ \\
$6.5$ & $7$ & $2$ & $5/7 = 0.71$ & $0.9\times 0.71 = 0.64$ \\
$10$ & $4$ & $1$ & $3/4 - 0.75$ & $0.64 \times 0.75 = 0.48$ \\
$12$ & $2$ & $1$ & $1/2= 0.5$ & $0.48 \times 0.5 = 0.24$ \\
$15$ & $1$ & $1$ & $0/1 = 0$ & $0$ \\
\end{tabular}
$$

## Kaplan-Meier Estimator and Log Rank test (Example)

* Patients who had survived a lobar intracerebral hemorrhage and whose genotype was known 
* The outcome variable was the time until recurrence of lobar intracerebral hemorrhage
* The investigators were interested in examining the genetic effect on recurrence as there were three common alleles e2, e3, and e4
* The analysis was focused on the effect of homozygous e3/e3 (Group 1) versus at least one e2 or e4 (Group 2)
* In the output are survival times (in months) for 71 subjects
* However, one subject’s genotype information is missing and is excluded from analysis
* Of the remaining 70 subjects, 32 are in Group 1 and 38 are in Group 2 
* A + sign indicates a censored observation; meaning that at that point in time the subject had yet to report recurrence

## Kaplan-Meier Estimator and Log Rank test (Example)

```{r out.width=".8\\linewidth"}
fit = with(hemorrhage, survfit(Surv(time,recur)~genotype))
plot(fit,lty=1:2, ylab='Probability of Hemorrhage-Free Survival', xlab='Time (in Months)',
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
legend('bottomleft',c('Group 1', 'Group 2'),lty=1:2,bty='n')
```

## Kaplan-Meier Estimator and Log Rank test (Example)

```{r}
with(hemorrhage, survdiff(Surv(time,recur)~genotype))
```

* Note that the log-rank test statistic is 6.3 with p-value 0.0122 based on a null $\chi^2$-distribution with 1 degree of freedom
* Thus the log-rank test confirms the difference in survival time of the two groups

## Gehan's Test

* Alternative to the log-rank test
* Generalization of the Wilcoxon procedure
* Randomized controlled trial subjects are randomized to one of two treatments, with survival times represented by $X$ and $Y$
* Let the sample as $X_1,\dots,X_{n1}$ and $Y_1,\dots,Y_{n2}$ with a censored observation denoted by $X_i^+$
* Only unambiguous pairs of observations are used
* Not used are ambiguous observations such as when an observed $X$ is greater than a censored $Y (X_i > Y_j^+)$ or when both observations are censored

## Gehan's Test

* The test statistic is defined as the number of times each of the $X$ clearly beats $Y$ minus the number of times $Y$ clearly beats $X$
* Let $S_1$ denote the set of uncensored observations
* Let $S_2$ denote the set of observations for which $X$ is censored and $Y$ is uncensored
* Let $S_3$ denote the set where $Y$ is censored and $X$ is uncensored
* Then Gehan's test statistic is
$$
\begin{aligned}
U = & \left( \#_{S_1} \{X_i > Y_j \} + \#_{S_2} \{X_i^+ \ge Y_j\} \right) \\
    & - \left( \#_{S_1} \{Y_j > X_i \} + \#_{S_3} \{ Y_j^+ \ge X_i \} \right)
\end{aligned}
$$

## Gehan's Test (Example)

* Experiment to assess the effect of a new treatment relative to a standard
* There are three required arguments to the function: 
    * the survival time, 
    * an indicator variable indicating that the survival time corresponds to an event (and is not censored), 
    * and a dichotomous variable representing one of two treatments
* Data:  
```S 94  180+ 741  1133 1261 382 567+ 988   1355+
   N 155 375  951+ 1198 175  521 683+ 1216+```

```{r echo=TRUE}
with(cancertrt,gehan.test(time,event,trt))
```

## Cox Proportional Hazards Models

* Let $x$ denote the corresponding $p \times 1$ vector of covariates
* Let $T_0$ denotea baseline response: a response in the absence of all covariate effects
* The hazard function of $T$ is the instantaneous chance of the event (death)
$$h(t) = \frac{f(t)}{S(t)}$$
* Assume that $T_0$ has the exponential distribution with pdf $f(t) = \lambda_0 \exp{-\lambda_0 t}, t > 0$
* Then the hazard function of $T_0$ has the constant value of $\lambda_0$
* The proportional hazards model assumes that the hazard function of $T$ is given by
$$\lambda(t;\boldsymbol{x}) = \lambda_0 e^{\boldsymbol{\beta}^T \boldsymbol{x}}$$
where $x$ is a $p \times 1$ vector of covariates and $\boldsymbol{\beta}$ is a $p \times 1$ vector of paramets
* Note that the hazard function of $T$ is proportional to that of $T_0$

## Cox Proportional Hazards Models

* To illustrate these ideas, assume that $T_0$ has constant hazard $\lambda_0$
* Suppose the only covariate is an indicator variable $w$ which is either 0 or 1 depending on whether a subject is not treated or treated
* Assuming a proportional hazards model, the hazard function of $T$ is given by
$$\lambda(t;w) = \lambda_0 e^{w\Delta}$$
* The hazard ratio of the experimental treatment relative to the control is then $e^{\Delta}$
* That is, $\Delta$ has the interpretation of log hazard
      * a value $< 1$ (less hazardous) favors the experimental treatment and 
      * a value $> 1$ (more hazardous) favors the control

## Cox Proportional Hazards Models
      
* The proportional hazards model developed by (Cox 1972) is a semiparametric model which does not necessarily specify the hazard function
* only the relative effect of covariates is estimated
* In the simple case under discussion it can be used to estimate the parameter $\Delta$ as shown in the following example

```{r}
fit = coxph(Surv(time,recur)~genotype,data=hemorrhage)
summary(fit)
```

## Cox Proportional Hazards Models

* More generally, assume that the baseline hazard function is $\lambda_0(t)$ 
* Assume that the hazard function of $T$ is
$$\lambda(t;\boldsymbol{x}) = \lambda_0(t) e^{ \boldsymbol{\beta}^T \boldsymbol{x} }$$
* Notice that the hazard ratio of two covariate patterns (e.g. for two subjects) is independent of baseline hazard
$$\frac{\lambda(t;\boldsymbol{x}_1)}{\lambda(t;\boldsymbol{x}_2)} = e^{\boldsymbol{\beta} (\boldsymbol{x}_1 - \boldsymbol{x}_2)} $$

## Cox Proportional Hazards Models (Example)

* Under investigation in this clinical trial was the pharmaceutical agent diethylstilbestrol DES 
* Subjects were assigned treatment to 1.0 mg DES (treatment = 2) or to placebo (treatment = 1)
* Covariates include age, serum hemoglobin level, size, and the Gleason index
* For demonstration purposes, we have dropped age and shb from the model

## Cox Proportional Hazards Models (Example)

```{r}
f2 = coxph(Surv(time,event=status)~as.factor(treatment)+size+index,data=prostate)
summary(f2)
```

## Cox Proportional Hazards Models (Example)

* These data suggest that the Gleason Index is a significant risk factor of mortality (p-value = 0.0356)
* Size of tumor is marginally significant (p-value = 0.0819)
* Given that $\widehat{\Delta} = -1.11272 < 1$ it appears that DES lowers risk of mortality
* However, the p-value = 0.3550 is nonsignificant


## References

* The Statistical Analysis of Failure Time Data (2002). Kalbfleisch and Prentice
* Analysis of Survival Data (1984). Cox and Oakes
* Ibrahim (2005). [Lecture Notes](http://www.amstat.org/chapters/northeasternillinois/pastevents/presentations/summer05_Ibrahim_J.pdf)
