---
title: "Time to Event Analysis (Part 2)"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  beamer_presentation:
    incremental: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
library(survival)
library(npsm)
library(boot)
```

## Overview

* Last time: 
    * Survival data
    * Survival function
    * Hazard function
    * Kaplan-Meier estimator

* Today:
    * Confidence bands with the Boostrap
    * Comparison of survival curves

## Confidence bands with the Boostrap

* Efron 1981 on Channing house data
```{r}
head(channing[channing$sex=="Male",])
```
* 97 men in retirement house in Palo Alto
* From opening 1964 until data collection day in July 1975
* 46 were observed exactly, the men died while in the Channing house
* The remaining 51 were censored, five moved elsewhere, and 46 were still alive at data collection day

## Confidence bands with the Boostrap

```{r out.width=".8\\linewidth"}
channingMale = channing[channing$sex=="Male",]
fit = survfit(Surv(time, cens) ~ sex, data = channingMale) 
plot(fit,ylab='Probability of Survival', xlab='Time (in Months)', main = 'Channing House Men',
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5, conf.int=FALSE)
abline(h = 0.5)
```

## Confidence bands with the Boostrap

* The quantify of interest was the median survival time
* Randomly censored data consist of iid pairs of observations $(X_i, \delta_i), i = 1,\delta,n$, 
    * if $\delta_i = 0$ then $X_i$ denotes a censored observation, and 
    * if $\delta_i = 1$ then $X_i$ denotes an exact "survival" time
* Efron takes a random sample with replacement from $(X_1,\delta_1),\dots,(X_n,\delta_n)$
* Then recomputes the survival function on the bootstrap sample
* There are theoretical developements by Akritas in 1986 showing that the boostrap approach works

## Log Rank Test

* We could use use our asymptotic results on the KM estimator $\widehat{S}(t)$ from last time or the bootstrap esimator to devise a test equality of survivar function at some time $t$
* But taking advantage of the entire function will make more efficient use of the data
* The most commonly used statistics are based on nonparametric rank tests

## Log Rank Test

* Let failure time $t_1 < \dots < t_k$ over both samples
* We construct contingency tables at every event time $t_i$
* $d_{0i}$ and $d_{1i}$ are the number of deaths in group 0 and 1
* $n_{0i}$ and $n_{1i}$ are the number at risk in groups 0 and 1
$$
\begin{tabular}{c|cc|c}
\hline
Group & Failures & Survivors & Total \\
\hline
$0$ & $d_{0i}$ & $n_{0i}-d_{0i}$ & $n_{0i}$ \\
$1$ & $d_{1i}$ & $n_{1i}-d_{1i}$ & $n_{1i}$ \\
\hline
Total & $d_{i}$ & $n_{i}-d_{i}$ & $n_{i}$ \\
\hline
\end{tabular}
$$
* Under the null hypothesis $S_1(t) = S_0(t), 0 < t < \infty$, $d_{1i}$ follows a hypergeometric distribution conditional on the margins

## Log Rank Test

* With the hypergeometric distribution we can get 
$$E_i = \operatorname{E}(d_{1i}) = n_{0i} \frac{d_i}{n_i}$$
$$V_i = \operatorname{Var}(d_{1i}) = \frac{\sum_{i=1}^k n_{1i} n_{0i}d_i(n_i-d_i)} {n_i^2(n_i-1)} $$
* And observed is $O_i = d_{0i}$
* Using this expectation and variances and summing over all timepoints $t_k$, we get the log rank statistics 
$$\chi^2 = \frac{\left(\sum_{i=1}^k (O_i - E_i)\right)^2}{\sum_{i=1}^k V_i}$$

## Log Rank Test

* Symmetric in two groups
* The log rank statistic depends on ranks of event times only
* $O-E$ from the $k$ two by two tables are treated as independent
* The $O-E$ in the numberator can be written as
$$\frac{d_{0i}d_{1i}}{d_i} (\widehat{\lambda}_{1i}-\widehat{\lambda}_{0i})$$
* So it quantifies the sums of differences in the hazard function over all $k$ time points scaled by by the fraction of failures

## Log Rank Test (Example)

* **Study**: Patients who had survived a lobar intracerebral hemorrhage and whose genotype was known 
* **Data**: Survival times (in months) for 71 subjects
* **Event**: Time until recurrence of lobar intracerebral hemorrhage
* **Question**: Genetic effect on recurrence in two gropus with different genotype
* One subject's genotype information is missing and is excluded from analysis
* Of the remaining 70 subjects, 32 are in Group 1 and 38 are in Group 2 
* A+ sign indicates a censored observation; meaning that at that point in time the subject had yet to report recurrence

## Log Rank Test (Example)

```{r out.width=".8\\linewidth"}
fit = with(hemorrhage, survfit(Surv(time,recur)~genotype))
plot(fit,lty=1:2, ylab='Probability of Hemorrhage-Free Survival', xlab='Time (in Months)',
     cex = 1.5, cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
legend('bottomleft',c('Group 1', 'Group 2'),lty=1:2,bty='n')
```

## Log Rank Test (Example)

```{r}
with(hemorrhage, survdiff(Surv(time,recur)~genotype))
```

* Note that the log-rank test statistic is 6.3 with p-value 0.0122 based on a null $\chi^2$-distribution with 1 degree of freedom
* Thus the log-rank test confirms the difference in survival time of the two groups

<!--
## Gehan's Test

* Alternative to the log-rank test
* Generalization of the Wilcoxon procedure
* Randomized controlled trial subjects are randomized to one of two treatments, with survival times represented by $X$ and $Y$
* Let the sample as $X_1,\dots,X_{n1}$ and $Y_1,\dots,Y_{n2}$ with a censored observation denoted by $X_i^+$
* Only unambiguous pairs of observations are used
* Not used are ambiguous observations such as when an observed $X$ is greater than a censored $Y (X_i > Y_j^+)$ or when both observations are censored

## Gehan's Test

* The test statistic is defined as the number of times each of the $X$ clearly beats $Y$ minus the number of times $Y$ clearly beats $X$
* Let $S_1$ denote the set of uncensored observations
* Let $S_2$ denote the set of observations for which $X$ is censored and $Y$ is uncensored
* Let $S_3$ denote the set where $Y$ is censored and $X$ is uncensored
* Then Gehan's test statistic is
$$
\begin{aligned}
U = & \left( \#_{S_1} \{X_i > Y_j \} + \#_{S_2} \{X_i^+ \ge Y_j\} \right) \\
    & - \left( \#_{S_1} \{Y_j > X_i \} + \#_{S_3} \{ Y_j^+ \ge X_i \} \right)
\end{aligned}
$$

## Gehan's Test (Example)

```{r echo=TRUE}
with(cancertrt,gehan.test(time,event,trt))
```
* Experiment to assess the effect of a new treatment relative to a standard
* There are three required arguments to the function: 
    * the survival time, 
    * an indicator variable indicating that the survival time corresponds to an event (and is not censored), 
    * and a dichotomous variable representing one of two treatments
* Data:  
```S 94  180+ 741  1133 1261 382 567+ 988   1355+
   N 155 375  951+ 1198 175  521 683+ 1216+```
-->

<!--
## Cox Proportional Hazards Models

* Let $x$ denote the corresponding $p \times 1$ vector of covariates
* Let $T_0$ denotea baseline response: a response in the absence of all covariate effects
* The hazard function of $T$ is the instantaneous chance of the event (death)
$$h(t) = \frac{f(t)}{S(t)}$$
* Assume that $T_0$ has the exponential distribution with pdf $f(t) = \lambda_0 \exp{-\lambda_0 t}, t > 0$
* Then the hazard function of $T_0$ has the constant value of $\lambda_0$
* The proportional hazards model assumes that the hazard function of $T$ is given by
$$\lambda(t;\boldsymbol{x}) = \lambda_0 e^{\boldsymbol{\beta}^T \boldsymbol{x}}$$
where $x$ is a $p \times 1$ vector of covariates and $\boldsymbol{\beta}$ is a $p \times 1$ vector of paramets
* Note that the hazard function of $T$ is proportional to that of $T_0$

## Cox Proportional Hazards Models

* To illustrate these ideas, assume that $T_0$ has constant hazard $\lambda_0$
* Suppose the only covariate is an indicator variable $w$ which is either 0 or 1 depending on whether a subject is not treated or treated
* Assuming a proportional hazards model, the hazard function of $T$ is given by
$$\lambda(t;w) = \lambda_0 e^{w\Delta}$$
* The hazard ratio of the experimental treatment relative to the control is then $e^{\Delta}$
* That is, $\Delta$ has the interpretation of log hazard
      * a value $< 1$ (less hazardous) favors the experimental treatment and 
      * a value $> 1$ (more hazardous) favors the control

## Cox Proportional Hazards Models
      
* The proportional hazards model developed by (Cox 1972) is a semiparametric model which does not necessarily specify the hazard function
* only the relative effect of covariates is estimated
* In the simple case under discussion it can be used to estimate the parameter $\Delta$ as shown in the following example

```{r}
fit = coxph(Surv(time,recur)~genotype,data=hemorrhage)
summary(fit)
```

## Cox Proportional Hazards Models

* More generally, assume that the baseline hazard function is $\lambda_0(t)$ 
* Assume that the hazard function of $T$ is
$$\lambda(t;\boldsymbol{x}) = \lambda_0(t) e^{ \boldsymbol{\beta}^T \boldsymbol{x} }$$
* Notice that the hazard ratio of two covariate patterns (e.g. for two subjects) is independent of baseline hazard
$$\frac{\lambda(t;\boldsymbol{x}_1)}{\lambda(t;\boldsymbol{x}_2)} = e^{\boldsymbol{\beta} (\boldsymbol{x}_1 - \boldsymbol{x}_2)} $$

## Cox Proportional Hazards Models (Example)

* Under investigation in this clinical trial was the pharmaceutical agent diethylstilbestrol DES 
* Subjects were assigned treatment to 1.0 mg DES (treatment = 2) or to placebo (treatment = 1)
* Covariates include age, serum hemoglobin level, size, and the Gleason index
* For demonstration purposes, we have dropped age and shb from the model

## Cox Proportional Hazards Models (Example)

```{r}
f2 = coxph(Surv(time,event=status)~as.factor(treatment)+size+index,data=prostate)
summary(f2)
```

## Cox Proportional Hazards Models (Example)

* These data suggest that the Gleason Index is a significant risk factor of mortality (p-value = 0.0356)
* Size of tumor is marginally significant (p-value = 0.0819)
* Given that $\widehat{\Delta} = -1.11272 < 1$ it appears that DES lowers risk of mortality
* However, the p-value = 0.3550 is nonsignificant
-->

## References

* The Statistical Analysis of Failure Time Data (2002). Kalbfleisch and Prentice
* [Lecture Notes (2005)](http://www.amstat.org/chapters/northeasternillinois/pastevents/presentations/summer05_Ibrahim_J.pdf). Ibrahim
* Efron (1981). Censored Data and the Bootstrap
* Akritas (1986). Bootstrapping the Kaplan-Meier Estimator
