---
title: "Homework 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

From our textbook (Ex. 2.8.17) [(Stanford library link)](https://searchworks.stanford.edu/view/10718635):  
In a large city, four candidates (Smith, Jones, Martinelli, and Wagner) are running for Mayor. A poll was conducted by random dialing with the following results: 
```{r}
candidate = c("Smith","Jones","Martinelli","Wagner","Others")
votes = c(442,208,460,180,205)
t(data.frame(candidate=candidate,votes=votes))
```
Using a 95% confidence interval, determine if there is a significant difference between the two front runners.

## Exercise 2

From our textbook (Ex. 2.8.18) [(Stanford library link)](https://searchworks.stanford.edu/view/10718635):  
In Example 2.7.1 we tested whether or not a dataset was drawn from a binomial distribution. For this exercise, generate a sample of size $n = 500$ from a truncated Poisson distribution as illustrated with the following R code:
```{r echo=TRUE,eval=FALSE}
x <- rpois(500,3)
x[x >= 8] = 7
```

(a) Obtain a plot of the histogram of the sample.
(b) Obtain an estimate of the sample proportion ``(phat<-mean(x/7))``.
(c) Test to see if the sample has a binomial distribution with $n = 7$, (i.e., use the same test as in Example 2.7.1).

## Exercise 3

From our textbook (Ex. 2.8.21) [(Stanford library link)](https://searchworks.stanford.edu/view/10718635):  
Even though the $\chi^2$-tests of homogeneity and independence are the same, they are based on different sampling schemes. The scheme for the test of independence is one-sample of bivariate data, while the scheme for the test of homogeneity consists of one-sample from each population. Let $C$ be a contingency table with $r$ rows and $c$ columns. Assume for the test of homogeneity that the rows contain the samples from the $r$ populations. Determine the (large sample) confidence intervals for each of the following parameters under both schemes, where $p_{ij}$ is the probability of cell $(i,j)$ occurring. Write R code to obtain these confidence intervals assuming the input is a contingency table.

1. $p_{11}$
2. $p_{11}−p_{12}$

## Exercise 4

Correspondence analysis is useful for text analysis. For instance, to compare texts from different authors in terms of common word frequency. The idea: Make a contingency table where rows are books from different authors and columns are words. 

Perform this type of analysis on the following two books form the ``languageR`` package. Use association plots, mosaic plots, and correspondence analysis, to determine if there are any words that contribute heavily to inhomogeneity among books. *Hint:* Use package ``vcd`` and ``ca``. Limit your analysis to words that occur frequently in both books.

```{r}
library(languageR)
data(alice,moby)
alice[1:20]
moby[1:20]
```

## Deadline

Please email your solutions in the form of a Rmd file to Nan Bi and Lexi Guan.

This homework is due on Monday, April 25th at 1:30 pm.

You are encouraged to work through this homework with your peers. But write your own code and explain the results in your own words.
