---
title: "Proportion Problem, Chi-Squared Test, and McNemar’s Test"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Previous Lectures

* One-sample tests: 
    * Sign test
    * Signed-Rank Wilcoxon
* Estimators, confidence intervals, and robustness to outliers
* Bootstrap
    * Error of estimators and significance for hypothesis testing
    * Complete enumerations
    * Tail proabilities

## Today

* One- and two-sample proportion problems
* $\chi^2$ Tests
* McNemar’s Test

## Proportion Problems

* Discrete variables
* The random variable $X$ consists of categories
* For now, we focus on binary categries failure (0) and success (1)
* So, we say $X \sim \operatorname{Bern}(p)$ is a random variable distributed according to the Bernoulli distribution 
    * with success probabiliy $p$ and 
    * failure probabililty $1-p$
* We know that $\operatorname{E}(X) = p$ and $\operatorname{Var} = p(1-p)$

## Proportion Problems

* Statistical problems can be 
    * estimating $p$,
    * forming confidence intervals,
    * and testing hypothesis 
    $$H_0: p = p_0 \text{ versus } H_A: p \ne p_0$$
* Let $X_1,\dots,X_n \sim \operatorname{Bern}(p)$ and $S$ be the total number of successes
* Then $S \sim B(p,n)$ follows a binomial distribution
* If $p$ is unkown but we assume that our observation come from a Bernoulli distribution, then
* We can estimate $p$ as $\widehat{p} = \frac{S}{n}$

## Proportion Problems

```{r echo=TRUE}
n = 10; p = 1/2; nsim = 10000; obsv = rbinom(nsim, size = n, prob = p)
```

```{r}
library(ggplot2)
df = data.frame(bern = table(obsv))
colnames(df) = c("S","Count")
ggplot(df, aes(x = S, y = Count)) + 
  geom_bar(stat="identity")
```

## Proportion Problems

```{r}
df <- data.frame(y = obsv)
ggplot(df, aes(sample = y)) + stat_qq() + ggtitle("Normal Q–Q Plot")
```

## Proportion Problems

```{r echo=TRUE}
n = 100; p = 1/2; nsim = 1000000; obsv = rbinom(nsim, size = n, prob = p)
```

```{r}
df = data.frame(bern = table(obsv))
colnames(df) = c("S","Count")
labels = floor(seq(0,n,length.out=25))
ggplot(df, aes(x = S, y = Count)) + 
  geom_bar(stat="identity") +
  scale_x_discrete(breaks=labels, labels=as.character(labels))
```

## Proportion Problems

```{r}
df <- data.frame(y = obsv)
ggplot(df, aes(sample = y)) + stat_qq() + ggtitle("Normal Q–Q Plot")
```

## Proportion Problems

* As $n \to \infty$ while $p$ is fixed:
    * The de Moivre–Laplace theorem (a special case of the central limit theorem) says $S$ approach a normal distribution
    * Easy confidence interval (just evaluate cdf of the normal)

## Example: Squeaky Hip Replacements

143 subjects with ceramic hip replacements

Ten report that their hip replacements squeaked
```{r echo=TRUE}
phat = 10/143
zcv = qnorm(0.975)
phat+c(-1,1)*zcv*sqrt(phat*(1-phat)/143)
```

We estimate between roughly 3 and 11% of patients who receive ceramic hip replacements will report squeaky replacements.

## Hypothesis Testing

* Test statistic is $z^2$ with
$$z = \frac{\widehat{p}-p_0}{\sqrt{p_0(1-p_0)/n}} \sim N(0,1)$$
* We are interested how large the observed proportion deviates this measured in terms of $|z|$
* Squared normal is distributed as $\chi^2$ distribution

## Example: Left-Handed Professional Ball Players

* Theory:  
    * Professional baseball players have different proportion of left handed player than left-handed people in genral population.
    * From previous study, we know that general public is has a proportion of $p = 0.15$.
* Hypothesis testing:
    * $H_0: p = 0 \text{ versus } H_A: p \ne 0.15$

## Example: Left-Handed Professional Ball Players

```{r warning=FALSE,echo=TRUE,message=FALSE}
library(Rfit)
head(baseball)
```

## Example: Left-Handed Professional Ball Players

```{r echo=TRUE}
ind = with(baseball,throw=='L')
n = length(ind)
phat = sum(ind)/n
phat
p0 = 0.15
z = (phat-p0)/(sqrt(p0*(1-p0)/n))
pvalue = 1-pchisq(z^2,df=1)
pvalue
```

## Why Not Use Exact Binomial Test?

* Since we know that $S$ follows a binomial distribution, why shouldn't we use it?
* Turns out that the binomial test is conservative

## $\chi^2$ Test

* Extension from two categories to multiple caterogries
* Consider discrete random variable $X$ with $1,2,\dots,c$ categories
* Let $p(j) = P(X = j)$ define the probabiliy mass function
* We wish to test:
$$H_0: p(j) = p_0(j), j = 1,\dots,c$$
$$H_A: p(j) \ne p_0(j), \text{ for some } j$$


