---
title: "Proportion Problems and Chi-Squared Tests"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: no
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
```

## Previous Lectures

* One-sample tests: 
    * Sign test
    * Signed-Rank Wilcoxon
* Estimators, confidence intervals, and robustness to outliers
* Bootstrap
    * Error of estimators and significance for hypothesis testing
    * Complete enumerations
    * Tail probability
* Observations from continuous distributions

## Today

* Observations from discrete distributions
* Proportion problems
* $\chi^2$ Tests

## Proportion Problems

* Discrete variables
* The random variable $X$ consists of categories
* For now, we focus on binary categories failure (0) and success (1)
* $X$ is a random variable distributed according to the Bernoulli distribution 
    * with success probabiliy $p$ and 
    * failure probabililty $1-p$
* We know that $\operatorname{E}(X) = p$ and $\operatorname{Var} = p(1-p)$

## Proportion Problems

* Statistical problems can be 
    * estimating $p$
    * forming confidence intervals around estimate $\widehat{p}$
    * and testing hypothesis 
    $$H_0: p = p_0 \text{ versus } H_A: p \ne p_0$$
* Let $X_1,\dots,X_n$ iid Bernoulli with success probability $p$ and $S$ be the total number of successes
* Then $S$ follows a binomial distribution with $n$ trials and success probability $p$
* If $p$ is unkown, we estimate $p$ with $\widehat{p} = \frac{S}{n}$

## Proportion Problems

```{r echo=TRUE}
n = 10; p = 1/2; nsim = 10000; obsv = rbinom(nsim, size = n, prob = p)
```

```{r}
library(ggplot2)
df = data.frame(bern = table(obsv))
colnames(df) = c("S","Count")
ggplot(df, aes(x = S, y = Count)) + 
  geom_bar(stat="identity")
```

## Proportion Problems

```{r}
df <- data.frame(y = obsv)
ggplot(df, aes(sample = y)) + stat_qq() + ggtitle("Normal Q–Q Plot")
```

## Proportion Problems

```{r echo=TRUE}
n = 100; p = 1/2; nsim = 10000; obsv = rbinom(nsim, size = n, prob = p)
```

```{r}
df = data.frame(bern = table(obsv))
colnames(df) = c("S","Count")
labels = floor(seq(0,n,length.out=25))
ggplot(df, aes(x = S, y = Count)) + 
  geom_bar(stat="identity") +
  scale_x_discrete(breaks=labels, labels=as.character(labels))
```

## Proportion Problems

```{r}
df <- data.frame(y = obsv)
ggplot(df, aes(sample = y)) + stat_qq() + ggtitle("Normal Q–Q Plot")
```

## Proportion Problems

* As $n \to \infty$ while $p$ is fixed:
    * The de Moivre–Laplace theorem (a special case of the central limit theorem) says $S$ approach a normal distribution
    * Easy confidence interval (just evaluate cdf of the normal)
    * Approximation of binomial distribution with $n$ trials and $p$ success by $N(np,np(1-p))$

## Example: Squeaky Hip Replacements {.build}

143 subjects with ceramic hip replacements

Ten report that their hip replacements squeaked

```{r echo=TRUE}
phat = 10/143
phat
zcv = qnorm(0.975)
phat+c(-1,1)*zcv*sqrt(phat*(1-phat)/143)
```

We estimate between roughly 3 and 11% of patients who receive ceramic hip replacements will report squeaky replacements

## Hypothesis Testing

* Reject null hypotheis if $|z|$ is large
$$z = \frac{\widehat{p}-p_0}{\sqrt{p_0(1-p_0)/n}} \sim N(0,1)$$
* $z$ is asymptotically standard normal 
* An equivalent test statistic of $|z|$ is $z^2$
* Squared normal is distributed as $\chi^2$ distribution

## Example: Left-Handed Professional Ball Players

* Theory:  
    * Professional baseball players have different proportion of left handed player than left-handed people in genral population
    * From previous study, we know that general public is has a proportion of $p_0 = 0.15$
* Hypothesis testing:
    * $H_0: p = 0.15 \text{ versus } H_A: p \ne 0.15$

## Example: Left-Handed Professional Ball Players

```{r warning=FALSE,echo=TRUE,message=FALSE}
library(Rfit)
head(baseball)
```

## Example: Left-Handed Professional Ball Players

```{r echo=TRUE}
ind = with(baseball,throw=='L')
n = length(ind)
phat = sum(ind)/n
phat
p0 = 0.15
z = (phat-p0)/(sqrt(p0*(1-p0)/n))
pvalue = 1-pchisq(z^2,df=1)
pvalue
```

## What is Nonparametrics Statistics Again?

<!--
* Rank-based methods: we do not estimate any additional parameters (exepct of course the parameter of interest $\theta$)
* Compare to $t$-test, we also have to estimate the variance 
* Tests for discrete observations: No parameter estimation neither in finite sample or asymptotic case
-->
* In all three nonparametric test (sign, signed-rank, $\chi^2$) no assumption on variances of observations
* In contrast, in the $t$-test variances are estimated

## Why Not Use Finite Sample Binomial Test?

Since we know that $S$ follows a binomial distribution, why don't we use it?

```{r echo=FALSE}
binom.test(sum(ind),n,p=p0)
```

* Finite sample $p$-values have upper bounded significance level $\alpha$
* Asymptotic $p$-values may be above above $\alpha$

## Why Not Use Finite Sample Binomial Test? {.build}

Problem is due to discreteness

Example: $n = 5$ and test $H_0: p = 0.5$ versus $H_A: p \ne 0.5$ 

Null distribution of $S$ is binomial with $n = 5$ and $p = 0.5$

Suppose outcome is $S = 5$ (most extreme observation)

```{r echo=TRUE}
n = 5; S = 5; p = 0.5
phat = S/n
pvalue = 2*p^S; pvalue
```

Problem is that the null hypethsis can never be true below $\alpha = 0.05$

So in this case, $\alpha$ has no meaning

## Discrete Random Variable (RV)

* Extension from two categories to multiple caterogries
* Consider discrete RV $X$ with $1,2,\dots,c$ categories
* Let $p(j) = P(X = j)$ define the probabiliy mass function
* We wish to test:
$$H_0: p(j) = p_0(j), j = 1,\dots,c$$
$$H_A: p(j) \ne p_0(j), \text{ for some } j$$

## Discrete RV

* $X_1,\dots,X_n$ is a random sample on $X$
* Let $O_j = \#\{ X_i = j \}$
* Observed frequencies are constrained $\sum_{j=1}^c O_j = n$ 
* The expected frequency for category $j$ is $\operatorname{E}_j = \operatorname{E}_{H_0}(O_j)$
* Two cases for $H_0$

## Discrete RV

* Case 1: 
    * All $p_0(j)$ are specified
    * So we get $E_j = np_0(j)$
    * Test stastitics is
    $$\chi^2 = \sum_{j=1}^c \frac{(O_j-E_j)^2}{E_j}$$
* Hypothesis $H_0$ is rejected in favor of $H_A$ for large values of $\chi^2$
* Observed frequencies, $(O_1,\dots,O_c)^T$ has a multinomial distribution, so exact distribution can be obtained
* Asymptotically $\chi^2$ distribution with $c−1$ degrees of freedom
* If we know $c-1$ frequencies, we can calculate the $c$th from total $n$

## Discrete RV Example {.build}

Roll a dice $n = 370$ times

Observe frequencies

```{r echo=TRUE}
O = c(58,55,62,68,66,61)
n = sum(O); n
```  

Test whether dice is fair $p(j) \equiv 1/6$

```{r echo=TRUE}
p0 = 1/6
E = rep(n*p0,6)
Chi2_0 = sum((O-E)^2/E); Chi2_0
```

## Discrete RV Example

Assymptotically equal to $\chi^2$ with $c-1$ degress of freedom

```{r echo=TRUE}
pvalue = 1-pchisq(Chi2_0,df=6-1); pvalue
```

Thus there is no evidence to support the dice being unfair

## Discrete RV

* Case 2: 
    * Only form of pmf is known
    * Have to estimate $p$
    * Same test stastitics but now with estimate $\widehat{p}$
    $$\chi^2 = \sum_{j=1}^c \frac{(O_j-E_j)^2}{E_j}$$
* Hypothesis $H_0$ is rejected in favor of $H_A$ for large values of $\chi^2$

## Discrete RV Example {.build}

Number of males in the first seven children for $n = 1334$ Swedish ministers of religion

```{r echo=TRUE}
males = 0:7
ministers = c(6,57,206,362,365,256,69,13)
n = sum(ministers); n
df = data.frame(ministers=ministers,males=males); t(df)
```

For example, 206 ministers had 2 sons and 5 daughters in their first 7 children

## Discrete RV Example {.build}

The maximum likelihood estimator of $p$ is

```{r echo=TRUE}
nChildren = n*7
nMale = sum(df$ministers*df$males)
phat = nMale/nChildren; phat
p0 = dbinom(males,7,phat)
E = n*p0
```

```{r}
df = data.frame(E=round(E,1),ministers=ministers,males=males)
t(df)
```

## Discrete RV Example {.build}

```{r echo=TRUE}
Chi2_0 = sum((df$ministers-E)^2/E)
pvalue = 1-pchisq(Chi2_0,df=8-1-1); pvalue
```

No evidence to refute a binomial probability model for the number of sons in the first seven children of a Swedish minister

## Discrete RV

* Alternatively to testing for deviation from a model
* We can get a confidence interval on pairwise difference in proportions $\widehat{p}_j - \widehat{p}_k$
* The confidence intervals are easy again because, we assume asymptotic normallity

## Discrete RV Example {.build}

Difference in the probabilities of all daughters or all sons

```{r}
t(df[,-1])
```

6 ministers had no sons, and 13 ministers had all sons

```{r echo=TRUE}
n = 1334; p0 = 6/n; p7 = 13/n
se = sqrt((p0+p7-(p0-p7)^2)/n)
zcv = qnorm(0.975)
lb = p0-p7 - zcv*se; ub = p0-p7 + zcv*se; res = c(p0-p7,lb,ub); res
```

Confidence interval covers 0, thus no significant difference in the proportions

## Several Discrete RVs

* Goal is to compare several discrete RV, which have same range $\{ 1,2,\dots,c \}$
* Consider hypothesis test:
    * $H_0:$ $X_1,\dots,X_r$ have the same distribution
    * $H_A:$ Distributions of $X_i$ and $X_j$ differ for some $i \ne j$
* Total number of samples $n = \sum_{i=1}^r n_i$
* Observed frequencies: 
$$O_{ij} = \#\{ \text{sample items in sample drawn on } X_i \text{ such that } X_i = j\},$$
* for $i = 1,\dots,r$ and $j = 1,\dots,c$
* $O_{ij}$ is a $r \times c$ matrix of observed frequency
* They are called *contingency tables*

## Several Discrete RVs

* Compare observed frequencies to the expected frequencies under $H_0$
* Estimate the common distribution $(p_1,\dots,p_c)^T$, where $p_j$ is the probability that category $j$ occurs
* Estimate probability of category $j$ overall
$$ \widehat{p}_j = \frac{\sum_{i=1}^r O_{ij}}{n}, j = 1,\dots,c$$
* Estimate expected frequencies $\widehat{E}_{ij} = n_i \widehat{p}_j$
* Notice that the sample size can vary between variables

## Several Discrete RVs

* Test statistics
$$\chi^2 = \sum_{i=1}^r \sum_{j=1}^c \frac{(O_{ij}-\widehat{E}_{ij})^2}{\widehat{E}_{ij}}$$
* Asymptotically $\chi^2$ with $(r−1)(c−1)$ degrees of freedom
* This is called *test for homogeneity*

## Several Discrete RVs Example {.build}

Distribution of alcoholic status same for different type of crime?

Contingency table with frequencies of  
criminals who committed crimes (6 RV's) and  
their alcoholics status (category: alcoholic and non-alcoholic)

```{r}
c1 = c(50,88,155,379,18,63) 
c2 = c(43,62,110,300,14,144)
ct = cbind(c1,c2)
colnames(ct) = c("Alcoholic","Non-Alcoholic")
rownames(ct) = c("Arson","Rape","Violence","Theft","Coining","Fraud")
ct
```

## Several Discrete RVs Example {.build}

```{r}
chifit = chisq.test(ct)
chifit
```

```{r echo=TRUE}
Chi2_0 = (chifit$observed-chifit$expected)^2/chifit$expected; Chi2_0
```

## Several Discrete RVs Example {.build}

Most of the contribution to the test statistic comes from the crime fraud

Eliminate fraud and retest

```{r}
chifit = chisq.test(ct[-6,])
chifit
```

Conclusion:  
Conditional on the criminal not committing fraud,  
cannot reject that alcoholic status has same distribution for all crimes

<!--
## McNemar's Test {.build}

Two discrete RV with two possible value $\{0,1\}$

$2 \times 2$ contingency table with four categories (0,0), (0,1), (1,0), and (1,1)

Denote probabilies of categories $p_{ij}$

Hypothesis test:
$$H_0: p_{01} - p_{10}$$
$$H_A: p_{01} \ne p_{10}$$

```{r}
ct = matrix(c("N_00","N_10","N_01","N_11"),ncol = 2)
rownames(ct) = colnames(ct) = c("0","1")
ct
```

## McNemar's Test

* The estimate of $p_{01}-p_{10}$ is 
$$\widehat{p}_{01} − \widehat{p}_{10} = (N_{01}/n) − (N_{10}/n)$$
same as difference in two proportions

## McNemar's Test Example

* Suppose A and B are two candidates for a political office who are having a debate
* Before and after the debate, the preference, A or B, of each member of the audience is recorded
* Given a change in preference of candidate, we are interested in the difference in the change from B to A minus the change from A to B
* If the estimate of this difference is significantly greater than 0, we might conclude that A won the debate
-->
