---
title: "Bootstrap (Example)"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output: 
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
  slidy_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question

Does small dose of aspirin prevent heart attacks in healthy middle-aged men?

## Experiment

* A randomized double-blind study was designed to collect data.   
* The subjects were randomly assigned to the aspirin and placebo groups.   
* Doctors and subjects didn't know whether they treated or received aspirin or placebo substance.   

```{r}
labels = c("nattacks","nsubjects")
aspirin = c(104,11037)
placebo = c(189,11034)
data = data.frame(aspirin,placebo)
rownames(data) = labels
data
```

## Statistic

Ratio of the rates:

```{r}
rate = function(v) v[1]/v[2]
theta.hat = rate(data$aspirin)/rate(data$placebo)
theta.hat
```

This means that in this sample aspirin-takers only have 55% as many heart attacks as placebo-takers.

## Computations

* Is this due to chance?   
* Or can this results be reproduced in a repetition of the same experiments?   
* Let's use the bootstrap method to find out:

```{r}
population.one = c( rep(1,data["nattacks","aspirin"]), 
                    rep(0,data["nsubjects","aspirin"]-data["nattacks","aspirin"]) )
population.two = c( rep(1,data["nattacks","placebo"]), 
                    rep(0,data["nsubjects","placebo"]-data["nattacks","placebo"]) )

draw.bootstrap.sample = function() {
  boot.pop.one = sample(population.one,replace = TRUE)
  boot.pop.two = sample(population.two,replace = TRUE)
  rate.one = sum(boot.pop.one)/length(boot.pop.one)
  rate.two = sum(boot.pop.two)/length(boot.pop.two)
  return(rate.one/rate.two)
}
```

## Computations

and now simulate:

```{r eval = FALSE}
nrep = 10000
theta.boot = replicate(nrep,draw.bootstrap.sample())
hist(theta.boot,breaks=100)

# sample estimate
abline(v=theta.hat,col = "red",lwd = 4)

# bootstrap confidence interval
confidence.lower = sort(theta.boot)[nrep*.025]
confidence.upper = sort(theta.boot)[nrep*.975]
abline(v=c(confidence.lower,confidence.upper),col = "blue",lwd = 4)
```

## Computations

```{r echo = FALSE}
nrep = 10000
theta.boot = replicate(nrep,draw.bootstrap.sample())
hist(theta.boot,breaks=100)

# sample estimate
abline(v=theta.hat,col = "red",lwd = 4)

# bootstrap confidence interval
confidence.lower = sort(theta.boot)[nrep*.025]
confidence.upper = sort(theta.boot)[nrep*.975]
abline(v=c(confidence.lower,confidence.upper),col = "blue",lwd = 4)
```

## Conclusion

We can conclude that aspirin was found to be significantly beneficial for preventing heart attacks.

