---
title: "Permutation Test"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
library(ggplot2)
library(HistData)
```

## The Lady and the Tea

* Fisher starts his second chapter in his famous book on "The Design of Experiments" first published in 1935:
* Quote: ``A lady declares that by tasting a cup of tea made with milk she can discriminate whether the milk or the tea infusion was first added to the cup.``
* Fisher proposed the the following experiment to test this
* Prepare eight cups of tea
* In four, pour the tea first
* In the other four, pour the milk first
* Hold constant other features of the cups of tea (size, temperature, etc.)
* Present cups in random order
* Lady tasts them and judges
* She knows that there are four of each type

## The Lady and the Tea

* Our theory is that the lady can taste the difference
* Which translates into our null hypothesis:  
$H_0:$ lady **cannot** taste difference
* The test statistics is the number of correct judgements
* Under the null all judgments are equally likely
* What is the null distribution of correct judgments?
* Once we have that compare it to the experiment and compute the probability of having observed the experimantal outcome or a more extreme case

## The Lady and the Tea

* Because of randomization, all $8! = 40320$ permutations of the cups are equally likely, and each one has its own number of correct judgements
* There are ${8 \choose 4} = 70$ ways of choosing 4 cups from 8 cups
* Think of it as, choosing 4 cups in succession by first choosing from 8 cups, then 7, 6, 5, which gives $8 \times 7 \times 6 \times 5 = 1680$
* But by doing so, we have chosen not only every possible set of 4, but every possible set in every possible order
* Since 4 cups can be rearranged in order in $4 \times 3 \times 2 \times 1 = 24$, we correct by $1680/24 = 70$

## The Lady and the Tea

```{r}
truth =     c("tea","milk","milk","milk","tea","tea","tea","milk")
judgement = c("tea","milk","milk","milk","tea","tea","milk","tea")
t(data.frame(truth = truth,judgement = judgement))
```

* Under the null, the reasons for the lady's judgements are unkown, except that they have nothing to do with the truth
* The judgements are what they are; they are fixed
* All $70$ ways to line up cups are equally likely
* But only one line up will match the lady's judgements

## The Lady and the Tea

```{r echo=TRUE}
t = 0:4
probability = (choose(4,t)*choose(4,4-t))/choose(8,4)
data.frame(t=t,probability=round(probability,digits=3))
```

* Of the four cups where the tea was poured first, select $t$ of them to say "tea" correctly, and $4−t$ to say "tea" incorrectly
* The chances of a match is $\frac{1}{70} = 0.014 < 0.05$
* We need perfect match!

## The Lady and the Tea

Monte Carlo simulations

```{r echo=TRUE}
judgement = c("tea","milk","milk","milk","tea","tea","milk","tea")
nSim = 10000
permutations = replicate(nSim,sample(8,replace = FALSE))
matches = apply(permutations,2,function(i) sum(judgement[i] == judgement))
data.frame(t=t,
           probability=round(probability,digits=3),
           monte=round(table(matches)/nSim,digits=3))
```

## The Lady and the Tea

```{r echo=TRUE}
t = 0:5
probability = (choose(5,t)*choose(5,5-t))/choose(10,5)
data.frame(t=t,probability=round(probability,digits=3))
```

* Increase experiment to total of 10 cups (5 each)
* If she tasted 10 cups, it would be possible to reject H0 without requiring perfect judgement
* We need perfect match or one miss will get us a $p$-value below $0.05$

## Darwin and Fisher

* In chapter 3 of Fisher's book, he then introduced the first nonparametric test developed
* He illustrated his approach on data from Darwin
* Darwin raised cross- and self-fertilized Zea mays plants
* He planted equal numbers of each in four different pots, but not same number in each pot
* Darwin measured heights of plants when only between 12 and 24 inches in height

## Darwin and Fisher

Original data (plant height in inches) from Darwin listed in Fisher's book

```{r }
data("ZeaMays")
ZeaMays
```

## Darwin and Fisher {.build}

Fisher proceeds by testing the hypothesis  
$H_0:$ No difference between crossed and self-fertilized plants

He peforms the paired sample $t$-test 
```{r}
t.test(ZeaMays$cross,ZeaMays$self,paired = TRUE,alternative = "two.sided")
```

## Darwin and Fisher {.build}

```{r echo=TRUE}
ZeaMays$diff
```
* Then he invents the first permutation test
* Under $H_0$ that self-fertilized versus cross-fertilized does not matter, only chance determined whether $A-B$ or $B-A$
* So the absolute value of the difference is what it is, but the plus or minus sign is by chance alone
* Test statistic is sum of the differences
* There are $2^{15} = 32768$ ways to swap the plus and minus signs, all equaly likely under $H_0$
* Calculate statistic for each one to get null distribution

## Darwin and Fisher

* Random samples $X_1,\dots,X_n$ from $X$
* Test $H_0: E(X) = 0$
* Write the usuall sample mean statistic as
$$T = \sum_{i=1}^n X_i = \sum_{i=1}^n |X_i| \delta_i$$
* and split samples into two parts
$$(|X_1|,\dots,|X_n|) \hspace{1cm}\text{and}\hspace{1cm} (\delta_1,\dots,\delta_n)$$
* Under $H_0$ the two parts are independent and $\delta_i$ is a fair coin flip $\{-1,1\}$
* The distribution of $T$ is not well defined under $H_0$

## Darwin and Fisher

* But If we fix $|X_i|$ at their observed values and regard $\delta_i$ as random
* Then the condition null of $T$ is well defined
* The only thing left to do is to compute
$$P\left(T \ge t \, \big| \, H_0, |X_1|, \dots, |X_n| \right)$$
by listing all possible sample of type
$$\left(\pm \, |X_1|, \dots, \pm \, |X_n|\right)$$
* This is usually called the 

## Darwin and Fisher

* We already looked at similar test
* The *sign test*: 
    * Test statistic: $S = \#_i \{ X_i \}$
    * Wilcoxon signed-rank test: $W = \sum_{i=1}^n \operatorname{R}(|X_i|) \delta_i$
* TODO: show why Wilcoxon signed-rank test is not conditional

## Darwin and Fisher

Using Monte Carlo simulations
```{r}

```

## General Recipe

* Decide on a test statistic $T$
* List the possible values of $T$
* Under $H_0$, all ways of re-arranging the data are equally likely
* $P(T = t)$ is proportional to the number of ways of getting the value $t$
* The permutation $p$-value is the probability of getting a value of $T$ as extreme or more extreme as the value we observed, "extreme" meaning in a direction inconsistent with $H_0$

## Assumption: Exchangeable Observations

* Book: Permutations, Parametric, and Boostrap Test of Hypothesis - Good, Section 2.2.2

## How to Compute It?

* Small sapmle size: Enumerations of all permutations (e.g. Gray codes)
* Medium sample size: Fast Frourier Transform
* Large sample size: Monte Carlo simulations, and Markov chain Monte Carlo

## Permutations and Fast Fourier Tranform

* Paper: On Obtaining Permutation Distributions in Polynomial Time

## More Advanced Example

* Paper: Equidistant Letter Sequences in the Book of Genesis

## Modern Application in Neuroimaging

* Explain neuroscience imaging data
* Explain significant cluster permutation test
* Paper: Nichols and Holmes, Nonparametric Permutation Tests For Functional Neuroimaging: A Primer with Examples

## Example Final Project: Autistic Brain

* Example for a project based on my autistic brain imaging example

## References

* Book:  Good. Permutations, Parametric, and Boostrap Test of Hypothesis
* Paper: Basu (1980). Randomization Analysis of Experimential Data: The Fisher Randomization Test
* Paper: On Obtaining Permutation Distributions in Polynomial Time
* Paper: Equidistant Letter Sequences in the Book of Genesis
* Paper: Nichols and Holmes, Nonparametric Permutation Tests For Functional Neuroimaging: A Primer with Examples
