---
title: "Nonlinear Regression"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
library(ggplot2)
library(ElemStatLearn)
```

## Overview

* Smoothing or estimating curves
    * Density Estimation
    * Nonlinear regression
* Rank-based linear regression

## Curve Estimation

* A curve of interest can be a probability density function $f$ or a regression function $r$
* In density estimation, we observe $X_1,\dots,X_n$ from some unknown cdf $F$ with density $f$
$$ X_1,\dots,X_n \sim f $$
and the goal is to estimate density $f$
* In regression, we observe pairs $(x_1,Y_1),\dots,(x_n,Y_n)$ that are related as
$$ Y_i = r(x_i) + e_i $$
with $E(e_i) = 0$, and the goal is to estimate the regression function $r$

## Density Estimation

```{r fig.width=10,fig.height=5}
dat <- data.frame(cond = factor(rep(c("A","B"), each=200)), 
                   rating = c(rnorm(200),rnorm(200, mean=.8)))
ggplot(dat, aes(x=rating)) + 
  geom_histogram(aes(y=..density..),binwidth=.1,colour="black", fill="white") +
  geom_density(alpha=.2, fill="#FF6666")
```

## Non-Linear Regression

```{r fig.width=10,fig.height=5}
data(bone)
ggplot(bone, aes(x=age, y=spnbmd, color=gender, shape=gender)) +  geom_point() +
  labs(title = "Bone Mineral Density Data") + labs(x = "Age") + labs(y = "Change in BMD")
```

## Non-Linear Regression

```{r fig.width=10,fig.height=5}
ggplot(bone, aes(x=age, y=spnbmd, color=gender, shape=gender)) +  geom_point() + geom_smooth() +
  labs(title = "Bone Mineral Density Data") + labs(x = "Age") + labs(y = "Change in BMD")
```

## The Bias–Variance Tradeoff



## References

* Wassermann (2006). All of Nonparametric Statistics
* Hollander, Wolfe, and Chicken (2014). Nonparametric Statistical Methods
