---
title: "Two-Sample Problems"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
```

## Two-Sample Problems

* Two populations
* From each population, we have one sample
* Infer whether or not there is a difference in location between populations
* Measure (with standard error) the difference (size of the effect) between populations
* Test: *Wilcoxon two-sample rank test*
* Estimation procedure: *Hodges–Lehmann estimate*

## Example

```{r message=FALSE,echo=TRUE}
library(HSAUR2); head(USmelanoma)
```

Data: Fisher and van Belle (1993) report mortality rates due to malignant melanoma of the skin for white males during the period 1950-1969, for each state on the US mainland

Mortality: Number of white males died due to malignant melanoma 1950-1969 per one million inhabitants

## Example

Question: Is there a difference in mortality for ocean states and non-ocean states?

```{r}
library(ggplot2)
ggplot(USmelanoma, aes(factor(ocean), mortality)) + geom_boxplot(width = 0.3)
```

## Stochastic Ordering of Alternatives

* Random sample $X_1,\dots,X_{n_1}$ with cdf $F$
* Random sample $Y_1,\dots,Y_{n_2}$ with cdf $G$
* Hypothesis test: $$H_0: F(t) = G(t) \hspace{1cm} \text{versus} \hspace{1cm} H_A: F(t) \le G(t)$$
* For the alternative, $X$ is stochastically larger than $Y$: $$P(X > t) \ge P(Y > t)$$

## Stochastic Ordering of Alternatives

```{r}
library(ggplot2)
library(reshape2)
t = seq(0,20,0.01)
df = data.frame(t=t,
                G=pgamma(t,shape = 9,scale = 0.5),
                F=pgamma(t,shape = 7.5,scale = 1))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
df = data.frame(t=t,
                g=dgamma(t,shape = 9,scale = 0.5),
                f=dgamma(t,shape = 7.5,scale = 1))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
t = seq(0,10,0.01)
df = data.frame(t=t,
                G=pnorm(t,mean = 3,sd = 0.7),
                F=pnorm(t,mean = 5,sd = 0.7))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
df = data.frame(t=t,
                g=dnorm(t,mean = 3,sd = 0.7),
                f=dnorm(t,mean = 5,sd = 0.7))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Test Based on Wilcoxon Scores

* Both samples are combined in one sample
* Combinded sample ranked from $1$ to $n_1+n_2$ (low to high)
* Denote $\operatorname{R}(Y_j)$ rank of observation $j$ in combined sample
* The Wilcoxon test statistics
$$ T = \sum_{j=1}^{n_2} \operatorname{R}(Y_j)$$
* $H_0$ is rejected for small values of $T$
* Under $H_0$ the two sample are from the same population:  
Any subset of ranks is equaliy likely as any other of the same size

## Test Based on Wilcoxon Scores

* For example, the probability that a subset of $n_2$ rankings is assigned to $Y$'s is $$\frac{1}{n_1+n_2 \choose n_2}$$
* There is no mention of the population distribution, so this test is *distriubiton free*
* So $p$-value can be calculated exactly by looking at the distribution of $T$ or aproximated asymptotically in case of large $n_1+n_2$

## Test Based on Wilcoxon Scores Example

Data: Case-control study of esophageal cancer in Ile-et-Vilaine, France (Breslow et al. 1980)

Hypothesis: Alcohol consumption same in the two groups

```{r}
library(datasets)
data(esoph)
head(esoph)
```

## Test Based on Wilcoxon Scores Example

```{r}
x<-rep(esoph$alcgp,esoph$ncases)
y<-rep(esoph$alcgp,esoph$ncontrols)
z<-c(x,y)
w<-c(rep(1,length(x)),rep(0,length(y)))
barplot(table(z,w),names.arg=c('Cases','Controls'), legend.text=levels(esoph$alcgp))
```

## Test Based on Wilcoxon Scores Example

```{r echo=TRUE}
x = as.numeric(x)
y = as.numeric(y)
wilcox.test(x,y)
```

Conclusion: Reject the null hypothesis in favor of the alternative that alcohol consumption is different in the two groups

