---
title: "Two-Sample Problems"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1234)
```

## Two-Sample Problems

* Two populations
* From each population, we have one sample
* Infer whether or not there is a difference in location between populations
* Measure (with standard error) the difference (size of the effect) between populations
* Test: *Wilcoxon two-sample rank test*
* Estimation procedure: *Hodges–Lehmann estimate*

## Example

```{r message=FALSE,echo=TRUE}
library(HSAUR2); head(USmelanoma)
```

Data: Fisher and van Belle (1993) report mortality rates due to malignant melanoma of the skin for white males during the period 1950-1969, for each state on the US mainland

Mortality: Number of white males died due to malignant melanoma 1950-1969 per one million inhabitants

## Example

Question: Is there a difference in mortality for ocean states and non-ocean states?

```{r}
library(ggplot2)
ggplot(USmelanoma, aes(factor(ocean), mortality)) + geom_boxplot(width = 0.3)
```

## Stochastic Ordering of Alternatives

* Random sample $X_1,\dots,X_{n_1}$ with cdf $F$
* Random sample $Y_1,\dots,Y_{n_2}$ with cdf $G$
* Hypothesis test: $$H_0: F(t) = G(t) \hspace{1cm} \text{versus} \hspace{1cm} H_A: F(t) \le G(t)$$
* For the alternative, $X$ is stochastically larger than $Y$: $$P(X > t) \ge P(Y > t)$$

## Stochastic Ordering of Alternatives

```{r}
library(ggplot2)
library(reshape2)
t = seq(0,20,0.01)
df = data.frame(t=t,
                G=pgamma(t,shape = 9,scale = 0.5),
                F=pgamma(t,shape = 7.5,scale = 1))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
df = data.frame(t=t,
                g=dgamma(t,shape = 9,scale = 0.5),
                f=dgamma(t,shape = 7.5,scale = 1))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
t = seq(0,10,0.01)
df = data.frame(t=t,
                G=pnorm(t,mean = 3,sd = 0.7),
                F=pnorm(t,mean = 5,sd = 0.7))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```

## Stochastic Ordering of Alternatives

```{r}
df = data.frame(t=t,
                g=dnorm(t,mean = 3,sd = 0.7),
                f=dnorm(t,mean = 5,sd = 0.7))
df = melt(df,id = "t")
ggplot(data = df,aes(x=t,y=value,group=variable,colour=variable)) + geom_line(size=1)
```


