---
title: "R and R Markdown Basics"
author: "Christof Seiler, STATS 205"
date: "Spring, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## For This Course

This class will require a lot of code and running computer simulations. 
All my examples and homeworks will be written in R. Choosing R for your homework solutions is highly recommended.

- R is an interpreted programming language, which means you will not have to compile your code. 
- R is very interactive, which means that you can play around with vectors and matrices, plot results.
- To keep track of your progress and to be able to construction your analysis steps, we will use R markdown. 
- R markdown is a format to make web reports that you can share with your collaborators.

## Vectors

Make a vector:

```{r}
x = c(11,218,123,36,1001)
y = rep(1,5)
z = seq(1,5,by=1)
x + y
z + 10
```

## Operations on Vectors

Some operations:

```{r}
sum(y)
c(mean(z),sd(z))
length(z)
```

## Coin Tossing

And most importantly there is a lot of functions for statistics. 

Like tossing a coin three times:

```{r}
set.seed(1)
coin = c('H','T')
samples = sample(coin,100,replace = TRUE)
```

Should come about half of the times head:

```{r}
sum(samples == 'H')
```

## Data Frames and Matrices


