---
title: "Robustness"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
  slidy_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Robustness of Estimators

* Robustness properties of the three estimators the mean, the median, and the Hodges–Lehmann estimate.
* Main measures of robustness:
    * efficiency, 
    * influence, 
    * and breakdown
* Today, we foucs on influence and breakdown

## Sensitivity Curve

* The finite sample version is the sensitivity curve
* The parameter $\theta$ is a function of the cdf $F$
* The observations $x_1,x_2,\dots,x_n$ are drawn from $F$
* Denote $\hat{\theta}$ as the estimator of $\theta$ in the sample
* Measure change in estimator $\hat{\theta}_n$ when outlier $x$ is added to a sample $x_n$
$$ x_n = ( x_1,x_2,\dots,x_n )^T $$
$$ x_{n+1} = ( x_1,x_2,\dots,x_n,x )^T $$
* The sensitivity curve of an estimator is
$$ S(x,\hat{\theta}) = \frac{\hat{\theta}_{n+1} - \hat{\theta}_n}{1/(n+1)} $$

## Sensitivity Curve Example

```{r echo=FALSE,warning=FALSE,message=FALSE}
library(ggplot2)
library(reshape2)
library(ICSNP)
```

```{r echo=TRUE}
x_n = c(1.85,2.35,-3.85,-5.25,-0.15,2.15,0.15,-0.25,-0.55,2.65)
mean(x_n)
median(x_n)
hl.loc(x_n)
```

## Sensitivity Curve Example

* Mean is unbounded
* Median and Hodges–Lehmann is bounded (estimators change slightly and changes soon become constant)

```{r fig.width=10,fig.height=4}
sensitivity = function(x,x_n,estimator) {
  (estimator(c(x_n,x))-estimator(x_n)) / (1/(length(x_n)+1))
}
interval = -20:20
SensData = data.frame(
  interval=interval,
  mean=sapply(interval,sensitivity,x_n,mean),
  median=sapply(interval,sensitivity,x_n,median),
  HL=sapply(interval,sensitivity,x_n,hl.loc)
)
SensDataLong <- melt(SensData, id = "interval")
ggplot(data = SensDataLong,
         aes(x = interval, y = value, colour = variable)) + 
         geom_line()
```

## Influence Function

* Sensitivity curve depends on sample $x_1,x_2,\dots,x_n$
* Influence function depends on the population distribution $F$
* It's *functional*: a function that takes another function as its argument
* An estimator is *robust* if its influence function is bounded
* The influence function for our estimators are (up to constant of proportionality and center)
    * Mean: $x$,
    * Median: $\sign(x)$, and 
    * HL: $F(x) − 0.5$
* Hence, mean is not robust, but median and HL are robust

## Influence Function Example


```{r fig.width=10}
library(gridExtra)
interval = seq(-3,3,0.1)
InfluenceData = data.frame(
  interval=interval,
  mean=interval,
  median=sign(interval),
  HL=pnorm(interval)-0.5
)
InfluenceDataLong <- melt(InfluenceData, id = "interval")
influencePlot = ggplot(data = InfluenceDataLong,
                       aes(x = interval, y = value, colour = variable)) + geom_line() + ggtitle("Influence Function")
sensativityPlot = ggplot(data = SensDataLong,
                         aes(x = interval, y = value, colour = variable)) + geom_line() + ggtitle("Sensitivity Curve")
grid.arrange(influencePlot, sensativityPlot, ncol=2)
```

## Breakdown Point of an Estimator



from http://www.stat.umn.edu/geyer/s06/5601/theo/break.pdf

## Relation of Ranks to Permutations

van der Vaart, Chapter 13
