---
title: "Two Sample Problems"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Autism Brain Imaging Data

We are going to look at a modern application where we encouter two sample problems very frequently. In neuroscience, we often compare two populations, such as people who are autistic and healthy controls, in the hope of finding morphological, connectivity, and functional difference between the two groups that could inform new therapies and cures.

Our running example will directly work on preprocessed neuroimaging data from the Autism Brain Imaging Data Exchange (ABIDE). 
The data is openly avaialbe on the ABIDE website [here](http://preprocessed-connectomes-project.org/abide/)).

ABIDE is a collaboration of 16 international imaging sites that have aggregated and are openly sharing neuroimaging data from 539 individuals suffering from ASD and 573 typical controls. For now will only use a subset of their data to speedup our analysis and foucs on 40 paricipants.

```{r}

```
