---
title: "Basic Statistics"
author: "Christof Seiler"
date: "Stanford University, Spring 2016, STATS 205"
output:
  ioslides_presentation:
    incremental: yes
    smaller: no
    transition: faster
    widescreen: yes
  beamer_presentation:
    incremental: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Sign Test

* Comparison of ice cream brands A and B
* Blindfolded taster gives preference of one ice cream over the other
* Experiment with 12 taster
    + 10 tasters prefer brand A over brand B
    + 1 taster prefers brand B over A
    + 1 no preference
* Pretty convincing evidence in favor of brand A
* But how likely is such a result due to chance if the null hypothesis is true (no preference in brands)?
* A *null hypothesis* a hypothesis associated with a contradiction to a theory one would like to prove

## Sign Test

* To every statistical test such as the sign test, we associate a *test statistic*
* A test *statistic* is a function of the data
* A *parameter* is a function of the probability distribution
* For this sign test, the test statistic $S$ is the number positive signs
* Evaluated on our data the observed test statistic is 10
* Under the null hypothesis $S$ is distributed according to a binomial distribution
* The binomial distribution is the number of success in a sequene of $n$ yes/no experiments with probability of success $p$

## Sign Test

A two-sidesd p-value for this problem can be calculated by evaluating the probability of obtaining having 10 success in 11 trials with success probability 1/2
```{r echo=FALSE}
2*dbinom(x=10,size=11,1/2)
```

```{r}

```

## Signed-Rank Wilcoxon Test


